Weighing Chances and Use Against Risks and Misuse of Emerging Genome Analysis and Manipulation Technology

### 2022-05-30 draft.tex
formatting

\author{\IEEEauthorblockN{1\textsuperscript{st} Benedict Bileam Scheuvens}
\IEEEauthorblockA{\textit{Data Science} \\
\textit{Hochschule Karlsruhe}\\
\textit{University of Applied Sciences} \\
scbe1047@h-ka.de}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Jenny Hilgenberg}
\IEEEauthorblockA{\textit{Data Science} \\
\textit{Hochschule Karlsruhe}\\
\textit{University of Applied Sciences} \\
Karlsruhe, Germany \\
hije1011@h-ka.de}
}

@@

\subsection{DNA and Identity}
"Identity" is a basic scientific concept in numerous disciplines, from psychology and pedagogy to sociology, ethnology, social and cultural anthropology, historical and literary studies to philosophy, and beyond that in many trans- and interdisciplinary debates, and has been for decades. Even if we do not ask ourselves who we are, we will receive unasked answers. Socialogist Abels describes identity as the answer to the question \textit{who am I} and therefore, identity as a differentiating term.\cite{abelsidentity}

Market leader Ancestry states to have over 20 million people in their database and therefore the world’s largest consumer DNA network\cite{aboutancestry}. What do people hope to gain from their DNA tests and what does it have to do with our identity? A great example for DNA information leading to a new self-understanding are "I am P’s father" (based on a paternity test) or "I am an African-American with my roots in Nigeria" (based on genetic ancestry tracing). In this case, identity is not only a matter of a unique narrative identity but also of numerical identity, i.e., identity in terms of individual continuity-over-time.\cite{nordgren}

@@

DNA tests are more popular than ever. Due to the decline in the cost of genetic analysis in the last two decades, "direct-to-consumer" genetic screening has become affordable for more and more people.\cite{rodney} In 2018, the global direct-to-consumer market was valued at approximately 824.1 million USD and is estimated to grow to 6,364.5 million USD by 2028. The rise in the public awareness along with the reduced cost and time required for sequencing owing to the technological advancements is responsible for bolstering the industry demand and aiding in reduction of the global healthcare expenditure.\cite{bisresearch}

@@

Many genetic testing companies appeal in their rhetoric to identity: The individualistic visions of identity is implicitly expressed in these quotations are clear examples of genetic individualism. Genetic individualism is the view that the DNA of an individual gives the individual a unique identity.\cite{nordgren} As mentioned in the chapter before, modern societies are searching for identity and this rhetoric is a response to that.\cite{nordgren} In addition to that, Nordgren and Juengst believe that the business models of these companies represent the confluence of three very different currents within contemporary Euro-American culture: the distinctly pre-modern search for a naturalistic understanding of individual identity in a pluralistic world, the thoroughly modern cachet of genomics as a science, and the post-modern emphasis on radical individual self-determination.

@@

Nevertheless, DNA should not be tracked and used lightly. However, statistics from the German Federal Criminal Police Office show that in addition to theft or murder, insults have also led to hits in DNA analysis\cite{bkastat}. What kind of cases make it legitimate to order a genome analysis after an insult is not explained – but the DNA profiles remain in the forensic database. In 2008, the European Court of Human Rights ruled that a retention regime that permits the indefinite retention of DNA records of both convicted and non-convicted (“innocent”) individuals is disproportionate\cite{smarpervsuk}. As a consequence, the Protection of Freedoms Act 2012 was implemented to permit the indefinite retention of DNA profiles of most convicted individuals and temporal retention for some first-time convicted minors and innocent individuals on the National DNA Database\cite{protectionoffreedomsact}. ??? MÖGLICHKEITEN ÜBER LÖSCHUNG DER DATEN IN ANDEREN LÄNDERN HERAUSSUCHEN ???

@@

Additionally there are numerous short term consequences. Preventing some but not all disease would result in amplification of marginalization, possibly increasing existing discrimination. \newline
The idea that disability is inherently wrong and needs to be fixed in itself is a questionable statement at best and strongly inhumane at worst\cite{schreiber}.

Assuming policies would be enacted on a national level instead of as a global initiative, practicing eugenics via genetic manipulation would deepen the divide between developed and developing nations causing additional inequality, alongside the social inequality unavoidibly caused by liberal eugenics, the approach of leaving the use of enhancements to choice.\newline
The opposite, obligatory genetic enhancement, would unearth even bigger concerns regarding bodily and reproductive autonomy\cite{eugenics}.

Artificially narrowing the human genepool might also have unforeseen sideeffects. As diversity is one of the driving agents of natural selection, careless intervention could inhibit our natural ability to defend against novel threats and disease\cite{mayr}.
It might quite literally be the end of evolution and the start of a new biological era. \newline
There are also outcries from a religious perspective arguing that taking responsibility away from god is divinely forbidden\cite{eugenics}.

>>>>

\author{\IEEEauthorblockN{1\textsuperscript{st} Benedict Bileam Scheuvens}
\IEEEauthorblockA{\textit{Data Science} \\
\textit{Hochschule Karlsruhe}\\
\textit{University of Applied Sciences} \\
Karlsruhe, Germany \\
scbe1047@h-ka.de}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Jenny Hilgenberg}
\IEEEauthorblockA{\textit{Data Science} \\
\textit{Hochschule Karlsruhe}\\
\textit{University of Applied Sciences} \\
Karlsruhe, Germany \\
hije1011@h-ka.de}
}

@@

\subsection{DNA and Identity}
"Identity" is a concept in numerous disciplines, from psychology and pedagogy to sociology, ethnology, social and cultural anthropology, historical and literary studies to philosophy, and beyond that in many trans- and interdisciplinary debates, and has been for decades. 
Even if we do not ask ourselves who we are, we will receive unprompted answers. Sociologist Abels describes identity as the answer to the question \textit{who am I} and therefore, identity as a differentiating term.\cite{abelsidentity}

Market leader Ancestry states to have over 20 million people in their database and therefore the world’s largest consumer DNA network\cite{aboutancestry}. What do people hope to gain from their DNA tests and what does it have to do with our identity? Great examples for DNA information leading to a new self-understanding are "I am P’s father" (based on a paternity test) or "I am an African-American with my roots in Nigeria" (based on genetic ancestry tracing). In this case, identity is not only a matter of a unique narrative identity but also of numerical identity, id est identity in terms of individual continuity-over-time.\cite{nordgren}

@@

DNA tests are more popular than ever. Due to the decline in the cost of genetic analysis in the last two decades, "direct-to-consumer" genetic screening has become affordable for more and more people\cite{rodney}. 
In 2018, the global direct-to-consumer market was valued at approximately 824.1 million USD and is estimated to grow to 6,364.5 million USD by 2028. 
The rise in the public awareness along with the reduced cost and time required for sequencing owing to the technological advancements is responsible for bolstering the industry demand and aiding in reduction of the global healthcare expenditure.\cite{bisresearch}

@@

Many genetic testing companies appeal in their rhetoric to identity: The individualistic visions of identity is implicitly expressed in these quotations are clear examples of genetic individualism. 
Genetic individualism is the view that the DNA of an individual gives the individual a unique identity\cite{nordgren}.
 As mentioned in the chapter before, modern societies are searching for identity and this rhetoric is a response to that\cite{nordgren}.
 In addition to that, Nordgren and Juengst believe that the business models of these companies represent the confluence of three very different currents within contemporary Euro-American culture: the distinctly pre-modern search for a naturalistic understanding of individual identity in a pluralistic world, the thoroughly modern cachet of genomics as a science, and the post-modern emphasis on radical individual self-determination.

@@

DNA should clearly not be tracked and used carelessly, however statistics from the German Federal Criminal Police Office show that in addition to theft or murder, cases of verbal abuse have also led to hits in DNA analysis\cite{bkastat}. 
What kind of cases make it legitimate to order a genome analysis after an insult is not explained – but the DNA profiles remain in the forensic database. In 2008, the European Court of Human Rights ruled that a retention regime that permits the indefinite retention of DNA records of both convicted and non-convicted (“innocent”) individuals is disproportionate\cite{smarpervsuk}. 
As a consequence, the Protection of Freedoms Act 2012 was implemented to permit the indefinite retention of DNA profiles of most convicted individuals and temporal retention for some first-time convicted minors and innocent individuals on the National DNA Database\cite{protectionoffreedomsact}. 
??? MÖGLICHKEITEN ÜBER LÖSCHUNG DER DATEN IN ANDEREN LÄNDERN HERAUSSUCHEN ???

@@

Additionally there are numerous short term consequences. Preventing some but not all disease would result in amplification of marginalization, possibly increasing existing discrimination.
The idea that disability is inherently wrong and needs to be fixed in itself is a questionable statement at best and strongly inhumane at worst, since it is deeply dehumanizing to talk about a disability problem in our society.
Yet this progression towards active filtering on which existence is allowed to happen is hardly avertable.
Outlawing prenatal diagnostics would only result in more late term miscarriages, suffering that could clearly have been minimized. No individual may be forced to carry a disabled fetus against their will either as their autonomy has to be respected just as much as the fetuses\cite{schreiber}.
This dilemma is not one that can simply be solved with legislature or technological advancements, rather it has to be approached from a standpoint of compassion.

Assuming policies would be enacted on a national level instead of as a global initiative, practicing eugenics via genetic manipulation would deepen the divide between developed and developing nations causing additional inequality, alongside the social inequality unavoidibly caused by liberal eugenics, the approach of leaving the use of enhancements to choice.
The opposite, obligatory genetic enhancement, would unearth even bigger concerns regarding bodily and reproductive autonomy\cite{eugenics}.

Artificially narrowing the human genepool might also have unforeseen sideeffects. As diversity is one of the driving agents of natural selection, careless intervention could inhibit our natural ability to defend against novel threats and disease\cite{mayr}.
It might quite literally be the end of evolution and the start of a new biological era.
There are also outcries from a religious perspective arguing that taking responsibility away from god is divinely forbidden\cite{eugenics}.
